---
format: 
 html:
    toc: true
    toc-depth: 3
---

::: section-card
## Data preparation

This section documents the construction of the analytical dataset, including population identifiers, taxonomic matching, phylogenetic filtering, moderator preparation, and random-effect structure checks.
:::

## Packeges

The R packages required for data preparation.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ape)
library(brms)
library(maps)
library(knitr)

options(mc.cores = parallel::detectCores())
```

## Import data

### The input files used in the analysis: the mixed-paternity dataset and the dated Clements phylogenetic tree.

```{r}
#| label: load-data
#| include: false

data_file <- "mixed_pat_final.csv"
tree_file <- "summary_dated_clements.nex"

mixed_pat_raw <- readr::read_csv(data_file, show_col_types = FALSE)
tree <- ape::read.nexus(tree_file)
```

This table summarises the main input files used in the analysis, including the size of the raw dataset and the number of tips and internal nodes in the phylogenetic tree.

```{r}
#| label: data-summary-table
#| echo: false
#| message: false
#| warning: false

data_summary <- tibble::tibble(
  Item = c("Raw dataset", "Phylogenetic tree"),
  Summary = c(
    paste0(nrow(mixed_pat_raw), " rows and ", ncol(mixed_pat_raw), " columns"),
    paste0(length(tree$tip.label), " tips and ", tree$Nnode, " internal nodes")
  )
)

knitr::kable(data_summary)
```

## Prepare data

### Population identifier

Population identity was defined using study identity and rounded geographic coordinates, with latitude and longitude rounded to two decimal places prior to constructing the population identifier.

```{r}

mixed_pat <- mixed_pat_raw %>%
  mutate(
    Lat_round = round(Lat, 2),
    Long_round = round(Long, 2),
    population_id = paste(Study_ID, Lat_round, Long_round, sep = "_"),
    population_id = as.factor(population_id)
  )
```

### Study and species formatting

The following step formats the main grouping variables used in the models and calculates the total number of broods by combining mixed- and non-mixed-brood counts.

```{r}
#| label: study-species-formatting

mixed_pat <- mixed_pat %>%
  mutate(
    Study_ID = as.factor(Study_ID),
    species_phylo_brms = as.factor(Clements_name),
    species_nonphylo = as.factor(Clements_name),

    n_mixed_broods = as.numeric(n_mixed_broods),
    n_non_mixed_broods = as.numeric(n_non_mixed_broods),
    n_total_broods = n_mixed_broods + n_non_mixed_broods
  )
```

### Phylogeny preparation

Species names were harmonised between the mixed-paternity dataset and the phylogenetic tree prior to pruning the tree and generating the phylogenetic covariance matrix used in the Bayesian models.

```{r}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false

species_data <- mixed_pat %>%
  filter(!is.na(species_phylo_brms)) %>%
  pull(species_phylo_brms) %>%
  as.character() %>%
  unique()

species_tree <- tree$tip.label

species_for_tree <- intersect(species_data, species_tree)

tree_analysis <- ape::keep.tip(tree, species_for_tree)

phylo_mat <- ape::vcv.phylo(tree_analysis, corr = TRUE)

phylo_mat <- phylo_mat[
  sort(rownames(phylo_mat)),
  sort(colnames(phylo_mat))
]

mixed_pat <- mixed_pat %>%
  filter(species_phylo_brms %in% rownames(phylo_mat)) %>%
  mutate(
    species_phylo_brms = droplevels(species_phylo_brms),
    species_nonphylo = droplevels(species_nonphylo)
  )
```

### Phylogeny - summary

```{r}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false

phylo_summary <- tibble::tibble(
  Item = c(
    "Species matched to phylogeny",
    "Tips in source tree",
    "Species retained after phylogenetic filtering"
  ),
  Value = c(
    length(species_for_tree),
    length(tree$tip.label),
    n_distinct(mixed_pat$species_phylo_brms)
  )
)

knitr::kable(phylo_summary)
```

### Moderator variable preparation

```{r}

min_n <- 5

mixed_pat <- mixed_pat %>%
  mutate(
    marker_type_model = as.factor(marker_type_model),
    nest_box_model = as.factor(nest_box_model),
    nest_type_model = as.factor(nest_type_model),

    breeding_system = as.factor(breeding_system),
    mono_poly_model = as.factor(mono_poly_model),
    parental_care = as.factor(parental_care),
    cooperative_breeding = as.factor(cooperative_breeding),

    migration = as.factor(migration),
    coloniality = as.factor(coloniality),

    CS_max = as.numeric(CS_max),
    body_mass = as.numeric(body_mass),
    study_duration = as.numeric(study_duration_num),

    Lat_round = round(as.numeric(Lat), 2),
    abs_lat = abs(Lat_round),

    CS_max_sc = as.numeric(scale(CS_max)),
    body_mass_sc = as.numeric(scale(body_mass)),
    study_duration_sc = as.numeric(scale(study_duration)),
    abs_lat_sc = as.numeric(scale(abs_lat))
  )
```

## Preparation of the database

### Database overall study

```{r}
#| label: data-overall-study

data_overall_study <- mixed_pat %>%
  filter(
    !is.na(n_mixed_broods),
    !is.na(n_non_mixed_broods),
    !is.na(n_total_broods),
    n_total_broods > 0,
    !is.na(Study_ID),
    !is.na(species_phylo_brms)
  ) %>%
  mutate(
    Study_ID = droplevels(Study_ID),
    species_nonphylo = droplevels(species_nonphylo),
    species_phylo_brms = droplevels(species_phylo_brms),
    population_id = droplevels(population_id)
  )
```

### Database single moderators and multivariable moderator model

```{r}

data_nest_box <- data_overall_study %>%
  filter(!is.na(nest_box_model)) %>%
  droplevels()


data_migration <- data_overall_study %>%
  filter(!is.na(migration)) %>%
  droplevels()

data_coloniality <- data_overall_study %>%
  filter(!is.na(coloniality)) %>%
  droplevels()


data_cooperative <- data_overall_study %>%
  filter(!is.na(cooperative_breeding)) %>%
  droplevels()

data_study_duration <- data_overall_study %>%
  filter(!is.na(study_duration_sc)) %>%
  droplevels()

data_clutch_size <- data_overall_study %>%
  filter(!is.na(CS_max_sc)) %>%
  droplevels()

data_body_mass <- data_overall_study %>%
  filter(!is.na(body_mass), body_mass > 0) %>%
  mutate(
    body_mass = as.numeric(body_mass),
    log_body_mass = log10(body_mass),
    log_body_mass_sc = as.numeric(scale(log_body_mass))
  ) %>%
  filter(
    is.finite(log_body_mass),
    is.finite(log_body_mass_sc)
  ) %>%
  droplevels()

data_abs_latitude <- data_overall_study %>%
  filter(!is.na(abs_lat_sc)) %>%
  droplevels()

data_marker_type <- data_overall_study %>%
  filter(!is.na(marker_type_model)) %>%
  droplevels()


data_nest_type <- data_overall_study %>%
  filter(!is.na(nest_type_model)) %>%
  droplevels()


valid_breeding_levels <- data_overall_study %>%
  count(breeding_system) %>%
  filter(!is.na(breeding_system), n >= 5) %>%
  pull(breeding_system)

data_breeding_system <- data_overall_study %>%
  filter(
    !is.na(breeding_system),
    breeding_system %in% valid_breeding_levels
  ) %>%
  droplevels()

data_mono_poly <- data_overall_study %>%
  filter(!is.na(mono_poly_model)) %>%
  droplevels()

data_parental_care <- data_overall_study %>%
  filter(!is.na(parental_care)) %>%
  droplevels()



data_full_brouwer_griffith <- data_overall_study %>%
  filter(
    !is.na(n_mixed_broods),
    !is.na(n_total_broods),
    n_total_broods > 0,
    !is.na(marker_type_model),
    !is.na(study_duration_sc),
    !is.na(breeding_system),
    breeding_system %in% valid_breeding_levels,
    !is.na(nest_type_model),
    !is.na(migration),
    !is.na(coloniality),
    !is.na(CS_max_sc),
    !is.na(body_mass),
    body_mass > 0,
    !is.na(abs_lat_sc)
  ) %>%
  mutate(
    log_body_mass = log10(body_mass),
    log_body_mass_sc = as.numeric(scale(log_body_mass)),
    marker_type_model = droplevels(marker_type_model),
    breeding_system = droplevels(breeding_system),
    nest_type_model = droplevels(nest_type_model),
    migration = droplevels(migration),
    coloniality = droplevels(coloniality),
    Study_ID = droplevels(Study_ID),
    species_nonphylo = droplevels(species_nonphylo),
    species_phylo_brms = droplevels(species_phylo_brms)
  ) %>%
  filter(is.finite(log_body_mass_sc)) %>%
  group_by(breeding_system) %>%
  filter(n() >= 5) %>%
  ungroup() %>%
  droplevels()

saveRDS(data_full_brouwer_griffith, "data/data_full_brouwer_griffith.rds")
```

## Overlap between nest type and nest-box use

Nest-box use and nest type describe related aspects of the nesting environment. We therefore checked their overlap before deciding which nest-related predictor to retain in the multivariable moderator model.

The overlap shows that nest-box use is concentrated in cavity-nesting records: among records with both variables available, 72.1% of nest-box records are classified as cavity nests, and 67.2% of cavity-nesting records are nest-box records. Because nest-box use and nest type therefore encode partly overlapping biological and sampling information, the multivariable moderator model retains nest type and excludes nest-box use.

```{r}
#| label: nest-type-nest-box-overlap
#| echo: false
#| message: false
#| warning: false

nest_overlap_summary <- data_overall_study %>%
  summarise(
    Total_records = n(),
    Records_with_nest_box = sum(!is.na(nest_box_model)),
    Records_with_nest_type = sum(!is.na(nest_type_model)),
    Records_with_both = sum(!is.na(nest_box_model) & !is.na(nest_type_model))
  )

knitr::kable(nest_overlap_summary)

nest_overlap_table <- data_overall_study %>%
  filter(!is.na(nest_box_model), !is.na(nest_type_model)) %>%
  count(nest_type_model, nest_box_model, name = "n") %>%
  group_by(nest_type_model) %>%
  mutate(
    Percent_within_nest_type = round(100 * n / sum(n), 1)
  ) %>%
  ungroup() %>%
  rename(
    Nest_type = nest_type_model,
    Nest_box = nest_box_model,
    Records = n
  )

knitr::kable(nest_overlap_table)

nest_box_type_distribution <- data_overall_study %>%
  filter(!is.na(nest_box_model), !is.na(nest_type_model)) %>%
  count(nest_box_model, nest_type_model, name = "n") %>%
  group_by(nest_box_model) %>%
  mutate(
    Percent_within_nest_box_status = round(100 * n / sum(n), 1)
  ) %>%
  ungroup() %>%
  rename(
    Nest_box = nest_box_model,
    Nest_type = nest_type_model,
    Records = n
  )

knitr::kable(nest_box_type_distribution)
```


## Relationship among candidate random-effects grouping factors

Before fitting the models, we evaluated the degree of overlap and nesting among candidate random-effects grouping factors, including species identity, study identity, and the pragmatic population identifier.

```{r}
#| label: random-effect-structure
#| echo: false
#| message: false
#| warning: false

count_one_match <- function(data, group_var, match_var, comparison_label) {
  data %>%
    distinct({{ group_var }}, {{ match_var }}) %>%
    count({{ group_var }}, name = "n_matches") %>%
    summarise(
      Comparison = comparison_label,
      Groups_with_only_one_match = sum(n_matches == 1),
      Total_groups = n(),
      Percent_only_one_match = round(mean(n_matches == 1) * 100, 1),
      .groups = "drop"
    )
}

random_effect_structure <- bind_rows(
  count_one_match(data_overall_study, species_nonphylo, population_id, "Populations per species"),
  count_one_match(data_overall_study, population_id, species_nonphylo, "Species per population"),
  count_one_match(data_overall_study, Study_ID, population_id, "Populations per reference"),
  count_one_match(data_overall_study, population_id, Study_ID, "References per population"),
  count_one_match(data_overall_study, species_nonphylo, Study_ID, "References per species"),
  count_one_match(data_overall_study, Study_ID, species_nonphylo, "Species per reference")
)

knitr::kable(random_effect_structure)
```









