---
format:
  html:
    toc: true
    toc-depth: 3
---

::: section-card
## Multivariable moderator model

This section documents the multivariable moderator phylogenetic beta-binomial model for mixed-paternity prevalence. Alternative classification schemes capture overlapping but not identical aspects of avian reproductive organisation, so they were evaluated as separate representations rather than included simultaneously in the multivariable moderator model. These alternatives included the breeding-system classification of Brouwer and Griffith (2019), a BIRDBASE-derived mono-/polygamy classification, parental-care classifications based on Cockburn (2006) and Griesser et al. (2017), and cooperative-breeding status from the Co-BreeD database.

The multivariable moderator model uses the Brouwer and Griffith breeding-system classification because it provides the most complete and interpretable representation for the multivariable model. Rare Brouwer and Griffith categories represented by fewer than five observations were excluded during data preparation. Nest type is retained, whereas nest-box use is excluded because it partially overlaps with nest type in biological and sampling information. Cooperative breeding is excluded because it strongly overlaps with the Brouwer and Griffith breeding-system classification.
:::

## Packages

The R packages required for preparing the model dataset and fitting the multivariable moderator model.

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-fold: true

library(tidyverse)
library(brms)
library(posterior)
library(bayesplot)
library(loo)

options(mc.cores = parallel::detectCores())

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}
```

## Load prepared objects

```{r}
#| label: load-full-model-data
#| message: false
#| warning: false

data_overall_study <- readRDS("data/data_overall_study.rds")
phylo_mat <- readRDS("data/phylo_mat.rds")
```

## Load multivariable moderator model dataset

The multivariable moderator model dataset is prepared in the data-preparation chapter and saved as `data/data_full_brouwer_griffith.rds`.

```{r}
#| label: load-full-model-dataset
#| message: false
#| warning: false

data_full_brouwer_griffith <- readRDS("data/data_full_brouwer_griffith.rds")
```

## Dataset size

```{r}
#| label: full-model-dataset-size
#| echo: false
#| message: false
#| warning: false

full_model_dataset_size <- tibble::tibble(
  Dataset = "Multivariable moderator model with Brouwer and Griffith breeding-system classification",
  Observations = nrow(data_full_brouwer_griffith),
  Species = n_distinct(data_full_brouwer_griffith$species_nonphylo),
  Studies = n_distinct(data_full_brouwer_griffith$Study_ID)
)

knitr::kable(full_model_dataset_size)
```

## Reference levels

Categorical coefficients in the multivariable moderator model are interpreted relative to the reference levels shown below. Continuous predictors are interpreted per one standard-deviation increase after scaling.

```{r}
#| label: full-model-reference-levels
#| echo: false
#| message: false
#| warning: false

reference_levels <- tibble::tribble(
  ~Predictor, ~Reference_level, ~Compared_levels,
  "Marker type", "DNA fingerprinting", "microsatellite; mixed / other; SNPs",
  "Breeding system", "cooperative breeder", "lekking; monogamy; no social bond; polyandry",
  "Nest type", "cavity", "ground; open; other enclosed",
  "Migration", "Strong migrant", "Weak migrant / partial resident",
  "Coloniality", "non-colonial", "colonial"
)

knitr::kable(reference_levels)
```

## Model specification

The model formula is shown below.

```{r}
#| label: full-model-formula
#| eval: false
#| echo: true

full_brouwer_griffith_formula <- bf(
  n_mixed_broods | trials(n_total_broods) ~
    marker_type_model +
    study_duration_sc +
    breeding_system +
    nest_type_model +
    migration +
    coloniality +
    CS_max_sc +
    log_body_mass_sc +
    abs_lat_sc +
    (1 | Study_ID) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo_brms, cov = phylo_mat))
)
```

## Fitting code


```{r}
#| label: fit-full-model
#| eval: false
#| echo: true

model_full_brouwer_griffith <- brm(
  formula = full_brouwer_griffith_formula,
  family = beta_binomial(link = "logit"),
  data = data_full_brouwer_griffith,
  data2 = list(phylo_mat = phylo_mat),
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(
    adapt_delta = 0.99,
    max_treedepth = 15
  ),
  seed = 123
)

saveRDS(model_full_brouwer_griffith, "models/model_full_BG.rds")
```

## Model summary


```{r}
#| label: full-model-summary
#| message: false
#| warning: false
#| code-fold: true

full_model_path <- first_existing(c(
  "models/model_full_BG.rds",
  "models/model_full_brouwer_griffith.rds"
))

if (!is.na(full_model_path)) {
  model_full_brouwer_griffith <- readRDS(full_model_path)
  summary(model_full_brouwer_griffith)
} else {
  knitr::kable(
    tibble::tibble(
      Status = "Multivariable moderator model not fitted yet",
      Expected_file = "models/model_full_BG.rds",
      Note = "Run the multivariable moderator model script to create the saved model object."
    )
  )
}
```

## Fixed-effect forest plot


```{r}
#| label: full-model-fixed-effects-forest
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 6.8
#| fig-cap: "Posterior fixed-effect estimates from the multivariable moderator model; horizontal intervals show 95% credible intervals on the log-odds scale. Categorical coefficients are contrasts against the following reference groups: marker type = DNA fingerprinting, breeding system = cooperative breeder, nest type = cavity, migration = strong migrant, and coloniality = non-colonial. Continuous predictors are interpreted per one standard-deviation increase after scaling. Effects to the right of zero indicate higher mixed-paternity prevalence relative to the reference group or scaled mean, whereas effects to the left of zero indicate lower prevalence."

forest_plot_file <- "Figures/model_diagnostics/full_model_fixed_effects_forest_plot.png"

if (file.exists(forest_plot_file)) {
  knitr::include_graphics(forest_plot_file)
} else {
  cat("Multivariable moderator model fixed-effect forest plot is not available yet. Run the figure code to save it first.")
}
```


## Key predicted effects


```{r}
#| label: full-model-key-effects
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7.2
#| fig-cap: "Key predicted effects from the multivariable moderator model. Points show posterior predicted prevalence for categorical moderators, lines show posterior predicted prevalence for continuous moderators, and intervals or ribbons show 95% credible intervals."

key_effects_file <- "Figures/model_diagnostics/full_model_key_effects.png"

if (file.exists(key_effects_file)) {
  knitr::include_graphics(key_effects_file)
} else {
  cat("Multivariable moderator model key-effects figure is not available yet. Run the key-effects plotting script to save it first.")
}
```


## Model diagnostics

```{r}
#| label: full-model-diagnostics
#| echo: false
#| message: false
#| warning: false

diagnostics_file <- first_existing(c(
  "models/diagnostics_model_full_BG.csv",
  "models/diagnostics_model_full_brouwer_griffith.csv"
))

if (!is.na(diagnostics_file)) {
  readr::read_csv(diagnostics_file, show_col_types = FALSE) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    knitr::kable()
} else if (!is.na(full_model_path)) {
  draw_diagnostics <- posterior::summarise_draws(
    posterior::as_draws_df(model_full_brouwer_griffith),
    "rhat",
    "ess_bulk",
    "ess_tail"
  )

  sampler_diagnostics <- brms::nuts_params(model_full_brouwer_griffith)

  diagnostic_table <- tibble(
    model = basename(full_model_path),
    max_rhat = max(draw_diagnostics$rhat, na.rm = TRUE),
    min_bulk_ess = min(draw_diagnostics$ess_bulk, na.rm = TRUE),
    min_tail_ess = min(draw_diagnostics$ess_tail, na.rm = TRUE),
    divergent_transitions = sampler_diagnostics %>%
      filter(Parameter == "divergent__") %>%
      summarise(n = sum(Value == 1)) %>%
      pull(n),
    treedepth_hits = sampler_diagnostics %>%
      filter(Parameter == "treedepth__") %>%
      summarise(n = sum(Value >= 15)) %>%
      pull(n)
  )

  knitr::kable(diagnostic_table, digits = 3)
} else {
  cat("Multivariable moderator model diagnostics are not available yet.")
}
```








